---
title: "Brain size viz"
author: "szymek drobniak"
date: "3/19/2021"
output: rmdformats::robobook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Packages

(Note: not all packages are needed for this segment of code, i.e., for visualisation)

```{r, message = F}
rm(list = ls())
# setwd

pacman::p_load(
  phylolm, ape, caper, geiger, phylopath,
  lme4, magrittr, dplyr, ggplot2, MuMIn,
  ggpubr, ggrepel, MCMCglmm, asreml, ggtree,
  ggtreeExtra, ggnewscale, RColorBrewer, plotly
)
```

## 2. Load data (using external scripts)

The below code loads the comparative brain data, associated moderators and performs robust PCA analysis of climatic variables. In the last step rename rows of the data to species IDs. The plots produced in the process are parts of the data summarising process, mostly summaries of PC components rotations. *Note: this step can be bypassed by loading the `.Rdata` file containing the image of workspace, point (4) below.*

```{r, message = F}
source(file = './analyses_new/analysis_reboot/data_load_plotting.R')
phylo_vcv <- vcv.phylo(birdtree.Jetz, corr = T)
# source('climatic.R')
rownames(birds) <- birds$tip_label
```

## 3. Clean and prepare final tree

The tree gets loaded via the `data_load_plotting.R` script, the below code removes one duplicated species (*Parus minor* which is treated as *Parus major* in Jetz et al. taxonomy), it also removes zero-length branches (introducing lengths below the tolerance threshold of ultrameric tree), and trimms the tree to only the species from the main dataset.

```{r}
birds <- birds[-794,]
tree_jetz_data <- birdtree.Jetz
head(sort(tree_jetz_data$edge.length)) # check branch lengths
tree_jetz_data$edge.length[tree_jetz_data$edge.length == 0] <- 1e-10

Ntip(tree_jetz_data)

# generate final tree
tree_final <- drop.tip(tree_jetz_data, setdiff(tree_jetz_data$tip.label, birds$tip_label))
```

## 4. Recreate the workspace without loading the above files

**If you want to avoid loading the above - just load this workspace to have all necessary data ready.**
```{r}
# load("vis.Rdata")
```

## 5. Plot the tree

Define colour palette to colour the terminal branches according to Prum et al.'s super-orders.
```{r, out.width = "800 px"}
# colour palette generation
set3 <- colorRampPalette((brewer.pal('Set3', n = 12)))

# assign taxonomic orders to colours + assign internal nodes (no data) to gray
colours <- c(setNames(set3(15), levels(birds$PrumOrder)), Internal = "gray50")
orders <- colours[c(as.character(birds[tree_final$tip.label, "PrumOrder"]),
            rep("Internal", 1174))]
```

Generate initial tree and label tree tips according to the developmental mode.

```{r, out.width = "800 px"}
# generate base tree
p <- ggtree(tree_final, layout = "circular", color = orders)

# add tree data and tip points
p <- p %<+% birds + new_scale("color") +
  scale_color_manual(values = c("cornflowerblue", "darkorange"),
  name = "Development mode") +
  geom_tippoint(aes(color = devo_mode), size = 0.5)
```

Add bar plots of circularly generated data depicting residual brain size and body weight. The data rings are offset relative to default positions to make them more aesthetically pleasing. Both `geom_fruit()` calls define the `x` axis, which becomes the *radial* axis on the circular layout. The problem with it is that it is almost impossible to twek/scale/increase size so that it looks good on a tree with this data density - I usually edit it afterwards in Illustrator to make it more visible and readable. Axes can be removed by deleting the `axis.params = ...` statements.
```{r, out.width = "800 px"}

# add bars representing residual brain size coloured by time of feeding and body weight
p1 <- p + geom_fruit(
  geom = geom_bar,
  mapping = aes(x = res_brain, fill = time_fed),
  axis.params = list(axis = "x"),
  stat = "identity", orientation = "y", offset = 0.15
) +
  geom_fruit(
    geom = geom_bar, mapping = aes(x = weight), axis.params = list(axis = "x"),
    stat = "identity", orientation = "y", offset = -0.01, fill = "gray74"
  ) +
  scale_fill_gradientn(
    colors = c("limegreen", "gold", "maroon2"),
    name = "Time fed [log days]"
  )

plot(p1)
# the code below saves the PDF
# ggsave(filename = "vis_tree_210226.pdf", device = 'pdf')
```

## 6. Multipanel figure of brain size correlates from models

Below code generates the multipanel plots of moderators. Each block defines a custom colour scale congruent with the tree development mode scale, modifies the ggplot2 theme to align it with Nature journals figure guidelines. The four plots are then bound together in one figure.

```{r, fig.height = 8}
text_size <- 15
line_size <- 0.5
tick_length <- unit(7, 'pt')
text_colour <- 'black'
plot1 <- birds %>%
  ggplot(aes(x = time_fed, y = res_brain, color = devo_mode)) +
  geom_point(size = 2, alpha = 0.3) +
  geom_smooth(method = 'lm') +
  scale_color_manual(values = c('#d55e00', '#3db7e9')) +
  labs(x = "Ln(time fed + 1)", y = "Residual of ln(brain size)", color = 'Developmental mode') +
  theme_classic() +
  theme(text = element_text(size = text_size, family = "Helvetica", colour = text_colour),
        axis.line = element_line(size = line_size, lineend = 'square'),
        axis.ticks = element_line(size = line_size),
        axis.ticks.length = tick_length,
        axis.text = element_text(colour = text_colour),
        legend.position = 'bottom')
#plot1

plot2 <- birds %>%
  ggplot(aes(x = clutch_size, y = res_brain, color = devo_mode)) +
  geom_point(size = 2, alpha = 0.3) +
  geom_smooth(method = 'lm') +
  scale_color_manual(values = c('#d55e00', '#3db7e9')) +
  labs(x = "Ln(clutch size)", y = "Residual of ln(brain size)", color = 'Developmental mode') +
  theme_classic() +
  theme(text = element_text(size = text_size, family = "Helvetica", colour = text_colour),
        axis.line = element_line(size = line_size, lineend = 'square'),
        axis.ticks = element_line(size = line_size),
        axis.ticks.length = tick_length,
        axis.text = element_text(colour = text_colour),
        legend.position = 'bottom')
#plot2

plot3 <- birds %>%
  ggplot(aes(x = move_max, y = res_brain)) +
  geom_boxplot(fill = 'gray80') +
  scale_x_discrete(limits = levels(birds$move_max), labels = c('Migratory', 'Sedentary')) + 
  labs(x = "Migratory status", y = "Residual of ln(brain size)", color = 'Developmental mode') +
  theme_classic() +
  theme(text = element_text(size = text_size, family = "Helvetica", colour = text_colour),
        axis.line = element_line(size = line_size, lineend = 'square'),
        axis.ticks = element_line(size = line_size),
        axis.ticks.length = tick_length,
        axis.text = element_text(colour = text_colour),
        legend.position = 'bottom')
#plot3

plot4 <- birds %>%
  ggplot(aes(x = foraging_3, y = res_brain)) +
  geom_boxplot(fill = 'gray80') +
  scale_x_discrete(limits = levels(birds$foraging_3), labels = c('Other', 'Arboreal')) + 
  labs(x = "Foraging substrate",y = "Residual of ln(brain size)", color = 'Developmental mode') +
  theme_classic() +
  theme(text = element_text(size = text_size, family = "Helvetica", colour = text_colour),
        axis.line = element_line(size = line_size, lineend = 'square'),
        axis.ticks = element_line(size = line_size),
        axis.ticks.length = tick_length,
        axis.text = element_text(colour = text_colour),
        legend.position = 'bottom')
#plot4

ggarrange(plot1, plot2, plot3, plot4, nrow = 2, ncol = 2, common.legend = T, legend = 'top',
          labels = "auto")
# ggsave(filename = "vis_gridplot_210225.pdf", device = 'pdf', scale = 0.8)
```

The below produces supplementary plot of food PCA - it works similarly to the above segment.

```{r, fig.height = 8}
plot5 <- birds %>%
  ggplot(aes(x = time_fed, y = res_brain, color = devo_mode_full)) +
  geom_point(size = 2, alpha = 0.3) +
  geom_smooth(method = 'lm') +
  scale_color_manual(values = c('#d55e00', '#3db7e9', '#f748a5', '#359b73')) +
  labs(x = "Ln(time fed + 1)", y = "Residual of ln(brain size)", color = 'Developmental mode') +
  theme_classic() +
  theme(text = element_text(size = text_size, family = "Helvetica"),
        axis.line = element_line(size = line_size, lineend = 'square'),
        axis.ticks = element_line(size = line_size),
        axis.ticks.length = tick_length,
        legend.position = 'bottom')
plot5
ggsave(filename = "vis_alldevomode.pdf", device = "pdf", scale = 0.8)


plot6.11 <- birds %>%
  ggplot(aes(x = food_energy, y = fibres)) +
  geom_point(size = 2, alpha = 0.3) +
  geom_smooth(method = 'loess') +
  labs(x = "Overall food energy content", y = "Fibres in food") +
  theme_classic() +
  theme(text = element_text(size = text_size, family = "Helvetica"),
        axis.line = element_line(size = line_size, lineend = 'square'),
        axis.ticks = element_line(size = line_size),
        axis.ticks.length = tick_length,
        legend.position = 'bottom')
#plot6.11

plot6.22 <- birds %>%
  ggplot(aes(x = food_energy, y = food_h_level)) +
  geom_point(size = 2, alpha = 0.3) +
  geom_smooth(method = 'loess') +
  labs(x = "Overall food energy content", y = "Food handling levels") +
  theme_classic() +
  theme(text = element_text(size = text_size, family = "Helvetica"),
        axis.line = element_line(size = line_size, lineend = 'square'),
        axis.ticks = element_line(size = line_size),
        axis.ticks.length = tick_length,
        legend.position = 'bottom')
#plot6.22

plot6.33 <- birds %>%
  ggplot(aes(x = food_h_level, y = fibres)) +
  geom_point(size = 2, alpha = 0.3) +
  geom_smooth(method = 'loess') +
  labs(x = "Food handling levels", y = "Fibres in food") +
  theme_classic() +
  theme(text = element_text(size = text_size, family = "Helvetica"),
        axis.line = element_line(size = line_size, lineend = 'square'),
        axis.ticks = element_line(size = line_size),
        axis.ticks.length = tick_length,
        legend.position = 'bottom')
#plot6.33

#ggarrange(plot6.11, plot6.33, plot6.22, nrow = 2, ncol = 2)

# below we modify plot adding PCA arrows
plot6.33 <- birds %>%
  ggplot(aes(x = food_h_level, y = fibres)) +
  geom_point(size = 2, alpha = 0.3) +
  geom_smooth(method = 'loess') +
  labs(x = "Food handling levels", y = "Fibres in food") +
  theme_classic() +
  theme(text = element_text(size = text_size, family = "Helvetica"),
        axis.line = element_line(size = line_size, lineend = 'square'),
        axis.ticks = element_line(size = line_size),
        axis.ticks.length = tick_length,
        legend.position = 'bottom')

arr1 <- c(cor(birds$food_energy, birds$food_energy_RC) * 0.05 * sqrt(nrow(birds)-1))
arr2 <- c(cor(birds$food_energy, birds$food_h_level_nofibres_RC) * 0.05 * sqrt(nrow(birds)-1))
arr4 <- c(cor(birds$fibres, birds$food_energy_RC) * 0.05 * sqrt(nrow(birds)-1))
arr3 <- c(cor(birds$fibres, birds$food_h_level_nofibres_RC) * 0.05 * sqrt(nrow(birds)-1))
arr5 <- c(cor(birds$food_h_level, birds$food_energy_RC) * 0.05 * sqrt(nrow(birds)-1))
arr6 <- c(cor(birds$food_h_level, birds$food_h_level_nofibres_RC) * 0.05 * sqrt(nrow(birds)-1))
plot6.44 <- birds %>%
  ggplot(aes(x = food_h_level_nofibres_RC, y = food_energy_RC)) +
  geom_point(size = 2, alpha = 0.3, colour = 'red') +
  geom_segment(aes(x=0, y=0, xend=arr1,yend=arr2), arrow = arrow(length = unit(10, 'pt'))) +
  geom_segment(aes(x=0, y=0, xend=arr3,yend=arr4), arrow = arrow(length = unit(10, 'pt'))) +
  geom_segment(aes(x=0, y=0, xend=arr5,yend=arr6), arrow = arrow(length = unit(10, 'pt'))) +
  labs(x = "PC1", y = "PC2") +
  geom_text(x = -2, y = 0.3, label = 'Fibres in food', hjust = 'left', lineheight = 1) +
  geom_text(x = 0.1, y = 2.15, label = 'Overall food energy \ncontent', hjust = 'left', lineheight = 1) +
  geom_text(x = 1.25, y = 0.85, label = 'Food handling \nlevels', hjust = 'left', lineheight = 1) +
  theme_classic() +
  theme(text = element_text(size = text_size, family = "Helvetica"),
        axis.line = element_line(size = line_size, lineend = 'square'),
        axis.ticks = element_line(size = line_size),
        axis.ticks.length = tick_length,
        legend.position = 'bottom')
#plot6.44

ggarrange(plot6.11, plot6.33, plot6.22, plot6.44, nrow = 2, ncol = 2, labels = 'auto')
#ggsave(filename = 'vis_foodpca_2.pdf', device = 'pdf', scale = 0.8)
```

## 7. Produce final effects plots...

With multi-tree estimates added. This plot uses linear model estimates from MCMCglmm and summary stats from multi-tree models pooling posteriors.

In the first step we prepare the effects table, group estimates by rearranging them and singling-out mass effects, and sorting to the desired order.

```{r, fig.dim = c(12, 10)}
effects <- read.table("effects_estimates.csv", sep = ";", head = T, stringsAsFactors = T)

labels_alloc <- rev(as.character(subset(effects, set == "parental provisioning")$term))
names(labels_alloc) <- labels_alloc
labels_alloc <- labels_alloc[-which(labels_alloc == "mass")]
labels_esoc <- rev(as.character(subset(effects, set == "ecosocial")$term))
names(labels_esoc) <- labels_esoc
labels_esoc <- labels_esoc[-which(labels_esoc == "mass")]
labels_comb <- rev(as.character(subset(effects, set == "combined")$term))
names(labels_comb) <- labels_comb
labels_comb <- labels_comb[-which(labels_comb == "mass")]

allcomb <- which(labels_comb %in% labels_alloc)
escomb <- which(labels_comb %in% labels_esoc)

final_labels <- rev(c("mass", rev(labels_alloc), rev(labels_esoc), rev(labels_comb[-c(allcomb, escomb)])))
names(final_labels) <- NULL
final_labels
length(final_labels)
```

The main plot is only defines base aesthetics mappings.
```{r, fig.dim = c(12, 10)}
p <- ggplot(data = subset(effects),
            aes(x = estimate, y = term, xmin = lower_CI, xmax = upper_CI))
```

Then we add points, error bars, each time defining custom colour and shape scales.
```{r, fig.dim = c(12, 10)}

p4 <- p + geom_vline(xintercept = 0, colour = 'gray90', linetype = 2) +
  geom_errorbarh(height = 0, lwd = 0.75) +
  geom_point(aes(fill = type), size = 2.5, colour = "black", pch = 21, stroke = 1) + theme_classic() +
  scale_y_discrete(limits = final_labels, expand = expansion(add = 0.8)) +
  scale_fill_manual(values = c("#00d302",  "#ff3cfe", "#ffac3b", "#00c2f9")) +
  theme(axis.line.y = element_blank(), panel.grid.major.y = element_line(size = 0.5),
        text = element_text(size = 15, colour = 'black'),
        axis.title.y = element_blank(),
        axis.text = element_text(colour = "black"),
        strip.background = element_rect(fill = "gray90", linetype = 0)) +
  xlab("Effect stimate") + labs(shape = "Fraction p < 0.05", fill = "Predictor set") +
  facet_grid(cols = vars(factor(set, levels = c("parental provisioning", "ecosocial", "combined")))) +
  geom_point(aes(x = estmulti, y = term, shape = f_cat, colour = type), size = 2,
             stroke = 1, position = position_nudge(y = -0.40)) +
  scale_shape_manual(values = c(20,3,8), labels = c("<5%", "5%-99%", ">99%")) +
  scale_colour_manual(values = c("#afff2a",  "#ff92fd", "#ffdc3d", "#7cfffa"), guide = F)

p4
# ggsave("vis_coefall_210302_2", device = "pdf", scale = 0.7)
```